Parameter Recovery Validation

### 2025-05-30 nbs/concepts/parameter-recovery-validation/index.qmd
fix: correct parameter-data correspondence in prior predictive checks
- Fix fundamental issue where data and parameters were sampled independently
- Ensure proper workflow: sample parameters first, then generate data using those parameters
- Update documentation to emphasize parameter-data correspondence requirement
- Update examples to demonstrate correct approach
- Add explanatory comments about why this order is critical for validation

```python
import torch
import pyro
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns
from pathlib import Path

# PyroVelocity imports
from pyrovelocity.models.modular.factory import create_piecewise_activation_model
# from pyrovelocity.plots.piecewise import (
#     plot_piecewise_phase_portrait,
#     plot_piecewise_trajectories,
#     plot_activation_timing,
# )
# from pyrovelocity.plots.predictive_checks import (
#     plot_prior_predictive_checks,
#     plot_posterior_predictive_checks,
#     plot_parameter_recovery_diagnostics,
# )

@@

Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:

```python
# Generate prior predictive samples using the generic interface
print("Performing prior predictive checks...")

@@

# Generate prior predictive samples (no samples argument = sample from prior)
prior_predictive_adata = model.generate_predictive_samples(
    num_cells=200,
    num_genes=10,
    num_samples=100,  # Generate 100 prior samples
    return_format="anndata"
)

# Extract parameter samples for validation
torch.manual_seed(1)
pyro.set_rng_seed(1)
prior_parameter_samples = model.sample_system_parameters(
    num_samples=1000,
    constrain_to_pattern=False,  # Use full prior ranges
    n_genes=10,
    n_cells=200
)

# Plot prior predictive checks
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_adata=prior_predictive_adata,
    prior_parameters=prior_parameter_samples,
    figsize=(15, 10)
)

@@

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_parameter_samples.items():
    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")

# Note: Pattern classification and other specialized validation methods
# will be implemented later as separate functionality
```

>>>>

```python
import torch
import pyro
import numpy as np
import matplotlib.pyplot as plt
from pathlib import Path
import scanpy as sc

# PyroVelocity imports
from pyrovelocity.models.modular.factory import create_piecewise_activation_model
from pyrovelocity.plots.predictive_checks import (
    plot_prior_predictive_checks,
    plot_posterior_predictive_checks,
)

@@

Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns.

**Important**: For meaningful prior predictive checks, we must ensure the plotted data corresponds to the plotted parameters. This requires sampling parameters first, then generating data using those specific parameters.

```python
# Generate prior predictive samples with proper parameter-data correspondence
print("Performing prior predictive checks...")

@@

# STEP 1: Sample parameters from the prior first
prior_parameter_samples = model.sample_system_parameters(
    num_samples=1000,
    constrain_to_pattern=False,  # Use full prior ranges
    n_genes=10,
    n_cells=200
)

# STEP 2: Generate data using a subset of those parameters for visualization
# Take first parameter sample for data generation
torch.manual_seed(1)
pyro.set_rng_seed(1)
single_param_sample = {key: value[:1] for key, value in prior_parameter_samples.items()}

prior_predictive_adata = model.generate_predictive_samples(
    num_cells=200,
    num_genes=10,
    samples=single_param_sample,  # Use specific parameters, not random sampling
    return_format="anndata"
)

# Add UMAP and clustering for visualization
sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
sc.tl.umap(prior_predictive_adata, random_state=42)
sc.tl.leiden(prior_predictive_adata, random_state=42)

# STEP 3: Plot with corresponding parameters and data
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_adata=prior_predictive_adata,        # Data generated from known parameters
    prior_parameters=prior_parameter_samples,  # Full parameter distribution
    figsize=(15, 10),
    save_path="reports/docs/prior_predictive",
)

@@

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_parameter_samples.items():
    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
```
